A thermodynamic view of dusty protoplanetary disks

### 2017-09-23 response.txt

We thank the referee for a prompt report, which helped to improve the
clarity of the manuscript. We have made revisions in bold, and further
explain them below.

Major points:

* The section 2 is interesting as it relates the work done by
  pressure, when not in phase with density, to the growth of
  oscillations. Although I understand that it is a link between the
  pure gas context and the dusty flow case (because the drag force
  acting on the gas can dephase density), I think the authors should
  approach the point later in the paper. In my opinion, it would be
  more adequate to approach it at the end of or after the actual
  section 3 (single fluid description of dusty gas). Then it is
  relevant to make the parallel with the pure gas dynamics. It is also
  a way to explain the analogy that appears in parag. 3.3 "The
  physical reason for this analogy..."

  We appreciate the referee’s suggestion that section 2 might make
  sense if moved.  With respect, our choice for the placement of this
  section is meant for readers who lack the technical expertise of
  the reviewer. Specifically, our goal is to summarize a main result
  and a key physical insight, and thereby to motivate the technical
  derivation of the dusty/adiabatic gas analogy. We added a new
  paragraph at the beginning of section 2 to state its purpose. 
  We hope section 2 motivates further reading and of course any reader
  can revisit  section 2 (section 4.4 refers back to it).

@@

  While the referee is certainly correct that the single fluid analogy
  needs to made, we have attempted to explain the concept in the
  introduction. In any event, we hope the referee will agree to some
  authorial prerogative in the order of presentation.

* In section 6 on the numerical results, in particular parag. 6.5 and
  6.6, I wonder which growth rates are obtained with different radial
  mode numbers. It would be an advantage to see other values as in Fig
  8. In 6.5 kx is 30, does the instability change with smaller, larger
  modes? It is worse in 6.6, where kx is very large, 1800. Is it
  relevant to have such a high radial mode, that could barely be
  resolved in non linear numerical simulations? Finally, in a general,
  I think that a more physical summary of the results of the section 6
  should be provided. It is almost limited now to a description of the
  results.

  For section 6.5 we have updated Fig. 14 to include cases with
  different kx. All growth rates remain the same order of
  magnitude. The effect of a radial variation in rhod/rhog is also
  similar between different kx. However, surface modes only appear at
  high kx (as found for pure gas VSI, Barker & Latter, 2017)

@@

* The section 7 'Discussion' is for the moment a list of common
  instabilities that could be triggered by the effect of the drag in
  the framework of the one-fluid approximation. Even if it is very
  interesting, it may be too speculative; I would not limit the
  discussion only to this. The authors may first discuss more about
  the limitations/assets of their approach based on the numerical
  results obtained in section 6, which will give arguments to the
  possible numerical implementation of the method (see parag. 7.3). In
  particular, what is the impact of the terminal velocity
  approximation? How much reliable the results and the accuracy of the
  approach are with a finite St number? For example, what do you
  expect for a typical value of 0.01-0.1 (dust or pebble), for which
  it is known that the terminal velocity is inappropriate?

  We have added a new section 7.1 on the application and limitations
  of the one-fluid framework.

  The terminal velocity approximation from the requires ts to be small
  compared to other timescales in the problem. Since 1/Omega is the
  typical `fast' timescale for disks, we generally require ts*Omega <<
  1, but the exact critical value is problem-dependent.

  Fig. 4 shows that the one-fluid framework gives accurate
  streaming instability growth rates in dust-rich disks even for
  ts*Omega = 0.1. However, the dust-poor `linB' mode in Table 1 (also
  with ts*Omega=0.1) is not accurately captured by the one-fluid framework. 
  This suggest that the terminal velocity approximation is not the
  only condition required for the one-fluid equations to give accurate
  results. In the case of the streaming instability, it seems that one
  also requires rhod > rhog.

### 2017-09-23 summary.tex
minor mods to response (grammar only) and summary (a bit of content)

For the VSI we find dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed where the dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}. 
Since the dust-induced buoyancy forces
increase away from the midplane, we find dust-loading can stabilize
`surface modes'  of the VSI, that would otherwise have the largest
growth rates. {\bf We also show that radial variations in
  $\rhod/\rhog$ effectively translates to a non-uniform radial temperature
  profile, and the resulting vertical shear can lead to instability,
  even if the true sound-speed is constant. 
}

>>>>

For the VSI we find dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed where the dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}. 
Since the dust-induced buoyancy forces
increase away from the midplane, we find dust-loading can stabilize
`surface modes'  of the VSI, that would otherwise have the largest
growth rates. {\bf We also show that radial variations in
  $\rhod/\rhog$ can trigger a type of VSI, even when the 
  usual sources of vertical shear -- vertical dust gradients and radial 
  temperature gradients -- are completely absent.  
  %effectively translates to a non-uniform radial temperature
  %profile, and the resulting vertical shear can lead to instability,
  %even if the true sound-speed is constant. 
}
